# Differential Expression Analysis Male vs Female

## Introduction

Differential expression analysis is a key step in the analysis of RNA-seq data. It is used to identify genes that are differentially expressed between two or more conditions. This analysis is important because it can help identify genes that are involved in the biological processes that are being studied. In this report, we will use three different methods to perform differential expression analysis: DESeq2, edgeR, and limma. We will then compare the results from these methods to identify genes that are consistently identified as differentially expressed.

```{r setup}
#| warning: FALSE
set.seed(42)
# load necessary packages
libraries <- c(
  "DESeq2", "edgeR", "ggplot2", "dplyr", "tximport", "limma", "ggVennDiagram", "cowplot",
  "openxlsx", "DT", "org.Mm.eg.db", "clusterProfiler", "RColorBrewer", "pheatmap", "ReactomePA", "msigdbr", "tidyr"
)
loaded <- lapply(libraries, require, character.only = TRUE)

comparison <- "bySexRA"

projectDir <- getwd()
resDir <- file.path(projectDir, "results", comparison)
plotDir <- file.path(resDir, "plots")

# source all files in the R directory
invisible(lapply(list.files(file.path(projectDir, "R"), full.names = TRUE), source))

padj.cutoff <- 0.05
lfc.cutoff <- 0.6
```

```{r create_directories}
#| warning: FALSE
#| echo: FALSE
#| include: FALSE
dir.create(file.path(resDir, "tables"), showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(resDir, "objects"), showWarnings = FALSE, recursive = TRUE)
dir.create(plotDir, showWarnings = FALSE, recursive = TRUE)
```

```{r initialize_datatable}
#| include: FALSE
# This is a hack to get the datatables to render in the html inside a tabset
htmltools::tagList(DT::datatable(cars, extensions = "Buttons", style = "bootstrap4"))
```

```{r load_tx2gene}
#| warning: FALSE
#| echo: false
tx2gene <- read.table(file.path(projectDir, "tx2gene.tsv"))
colnames(tx2gene) <- c("TXNAME", "GENEID", "GENESYMBOL")
```

```{r setMetadata}
#| warning: FALSE
ddsDir <- file.path(resDir, "objects", "dds.rds")
if (file.exists(ddsDir)) {
  dds <- readRDS(ddsDir)
} else {
  raw_count <- read.table(file.path(projectDir, "raw_data", "GSE267573_raw_counts.csv"), header = TRUE, row.names = 1, sep = ",")
  # create a dataframe with a column named Samples with the column names of the raw counts
  metadata <- data.frame(Sample = colnames(raw_count)) |>
    dplyr::mutate(
      Condition = ifelse(grepl("H", Sample), "Oxygen", "Room_Air"),
      Sex = ifelse(grepl("F", Sample), "Female", "Male"),
      Condition = factor(Condition, levels = c("Room_Air", "Oxygen")),
      Sex = factor(Sex, levels = c("Female", "Male"))
    )
  rownames(metadata) <- metadata$Sample

  se <- CreateSummarizedExperiment(raw_count, metadata)
  # subset SummarizedExperiment to only include samples with Condition of Rooom_Air
  se <- se[, se$Condition == "Room_Air"]
  # Create a DESeqDataSet object
  dds <- DESeqDataSet(se, design = ~Sex)
  ddsTxi <- Filterdds(dds, groupSize = 2)
  dds <- DESeq(ddsTxi)
  saveRDS(dds, ddsDir)
}
```

### Heatmap of the sample-to-sample distances

```{r sample_to_sample_heatmap}
#| warning: FALSE
#| fig-width: 8
#| fig-height: 6
#| echo: false
sampleDistanceHeatmap(dds, labelColumns = c("Sex"))
```

### PCA plot

```{r pca_plot}
#| warning: FALSE
#| fig-width: 8
#| fig-height: 6
#| echo: false
pcaPlot(dds, intgroup = c("Sex"))
```

Voom is a transformation method that is used to transform count data into log-counts per million (logCPM) values. This transformation is used to stabilize the variance of the data, which is important for the downstream analysis when using limma.

### Differential Expression

```{r DESeq2_DE}
#| warning: FALSE
dgeDir <- file.path(resDir, "objects", "dge.rds")
if (!file.exists(dgeDir)) {
  dgeObj <- GetDGEObject(dds, group = "Sex")
  saveRDS(dgeObj, dgeDir)
}
dgeObj <- readRDS(dgeDir)

deResDir <- file.path(resDir, "objects", "DE.rds")
if (!file.exists(deResDir)) {
  DE <- GetDE(
    dds = dds, dge = dgeObj, comparisonNameDESeq2 = "Sex_Male_vs_Female", comparisonNameDGE = "SexMale",
    designDGE = ~Sex, alpha = padj.cutoff, foldChangeThreshold = lfc.cutoff, verbose = FALSE
  ) 
  saveRDS(DE, deResDir)
}
DE <- readRDS(deResDir)
```

::: {.callout-note collapse="true"}
## DESeq2 Pvalue Distribution
```{r pvalue_distribution}
#| warning: FALSE
#| echo: FALSE
# Histogram of p-value distribution
pvals <- DE$DESeq2$pvalue
hist(pvals, breaks = 50, col = "skyblue", border = "white", main = "Histogram of p-values", xlab = "p-value")
```
:::

## Results

Different methods for differential expression analysis can yield different results. The following Venn diagrams show the overlap between the significant genes identified by DESeq2, edgeR, and limma.

```{r venn}
#| warning: FALSE
#| fig-width: 16
#| fig-height: 6
#| echo: false
v1 <- ggVennDiagram::ggVennDiagram(
  list(
    DESeq2 = dplyr::filter(DE$DESeq2, shrunken_sig_exp %in% c("UP", "DOWN")) |> rownames(),
    edgeR = dplyr::filter(DE$edgeR, sig_exp %in% c("UP", "DOWN")) |> rownames(),
    limma = dplyr::filter(DE$limma, sig_exp %in% c("UP", "DOWN")) |> rownames()
  )
) + ggtitle("DESeq2, edgeR, and limma significant genes")

v2 <- ggVennDiagram::ggVennDiagram(
  list(
    DESeq2 = dplyr::filter(DE$DESeq2, shrunken_sig_exp == "UP") |> rownames(),
    edgeR = dplyr::filter(DE$edgeR, sig_exp == "UP") |> rownames(),
    limma = dplyr::filter(DE$limma, sig_exp == "UP") |> rownames()
  )
) + ggtitle("DESeq2, edgeR, and limma upregulated genes")

v3 <- ggVennDiagram::ggVennDiagram(
  list(
    DESeq2 = dplyr::filter(DE$DESeq2, shrunken_sig_exp == "DOWN") |> rownames(),
    edgeR = dplyr::filter(DE$edgeR, sig_exp == "DOWN") |> rownames(),
    limma = dplyr::filter(DE$limma, sig_exp == "DOWN") |> rownames()
  )
) + ggtitle("DESeq2, edgeR, and limma downregulated genes")

# plot the venn diagrams side by side
venns <- cowplot::plot_grid(v1, v2, v3, ncol = 3)
ggsave(file.path(plotDir, "venn_diagrams_methods.pdf"), venns, width = 16, height = 6)
print(venns)
```

The following table shows the significant genes identified by each method.
Download the tables by clicking the buttons above each table.

```{r DE_tables}
#| warning: false
#| panel: tabset
#| output: asis
#| echo: false
cat("## DESeq2\n")
# Create a table for the significant genes identified by each method
print(getDESeq2Table(DE$DESeq2))
cat("\n")
cat("## edgeR\n")
print(getEdgeRTable(DE$edgeR))
cat("\n")
cat("## limma\n")
print(getLimmaTable(DE$limma))
```

```{r volcanos}
#| warning: FALSE
#| echo: false
p <- volcanoPlot(DE$DESeq2, fc = "shrunken_log2FoldChange", padj = "padj", sig_exp = "shrunken_sig_exp", title = "DESeq2 Male vs Female", color = list(UP = "#96CECE", DOWN = "#94609F", NO = "grey"), maxPvalue = 18, maxLogFC = 3)
print(p)
ggsave(file.path(plotDir, "DESeq2_volcano_plot.pdf"), p, width = 6, height = 6)
```

```{r heatmap}
#| warning: FALSE
#| fig-width: 14
#| fig-height: 8
#| echo: false
ht1 <- heatmapPlot(
  dds,
  DE$DESeq2 |> dplyr::filter(shrunken_sig_exp %in% c("UP", "DOWN")),
  title = "DESeq2 Male vs Female",
  tx2gene
) 
ComplexHeatmap::draw(ht1, auto_adjust = FALSE)
pdf(file.path(plotDir, "DESeq2_heatmap.pdf"), width = 14, height = 8)
ComplexHeatmap::draw(ht1, auto_adjust = FALSE)
dev.off()
```

## Pathway Analysis

The genes identified as differentially expressed can be used to perform pathway analysis. This analysis can help identify the biological processes that are affected by the treatment. The following table shows the pathways that are enriched among the differentially expressed genes.

[Cluster Profiler - Overview of enrichment analysis](https://yulab-smu.top/biomedical-knowledge-mining-book/enrichment-overview.html)

```{r pathway_analysis}
#| warning: FALSE
#| echo: FALSE
pathwayListGSEDir <- file.path(resDir, "objects", "pathwayListGSE.rds")
if (!file.exists(pathwayListGSEDir)) {
  pathwaysGSE <- GetGSEA(DE$DESeq2, statCol = "stat", organism = "mouse", sigExpCol = "shrunken_sig_exp")
  saveRDS(pathwaysGSE, pathwayListGSEDir)
} 
pathwaysGSE <- readRDS(pathwayListGSEDir)

pathwayListOverRepDir <- file.path(resDir, "objects", "pathwayListOverRep.rds")
if (!file.exists(pathwayListOverRepDir)) {
  pathwaysOverRep <- GetOverRepAnalysis(DE$DESeq2, statCol = "stat", organism = "mouse", sigExpCol = "shrunken_sig_exp")
  saveRDS(pathwaysOverRep, pathwayListOverRepDir)
}
pathwaysOverRep <- readRDS(pathwayListOverRepDir)
```

```{r pathway_tables}
#| warning: FALSE
#| echo: FALSE
#| output: asis
#| panel: tabset
nameMap <- list(gse = "Gene Set Enrichment", overRep = "Over-representation")
descriptionMap <- list(
  gse = "Gene Set Enrichment Analysis (GSEA) is a computational method that determines whether an a priori defined set of genes shows statistically significant, concordant differences between two biological states.",
  overRep = "Over Representation Analysis (ORA) (Boyle et al. 2004) is a widely used approach to determine whether known biological functions or processes are over-represented (= enriched) in an experimentally-derived gene list, e.g. a list of differentially expressed genes (DEGs)."
)
cat(paste0("\n## ", nameMap$gse, "\n"))
cat(descriptionMap$gse, "\n")
cat("\n All genes as input:\n")
allGenes <- pathwaysGSE$summary |> dplyr::filter(Input == "all")
if (nrow(allGenes) == 0) {
  data <- data.frame()
} else {
  data <- table(allGenes$DB) |>
    as.data.frame() |>
    tidyr::pivot_wider(names_from = Var1, values_from = Freq)
}
print(htmltools::tagList(DT::datatable(data, rownames = FALSE, style = "bootstrap4", options = list(dom = 't'))))
cat("\n Only Significant genes as input:\n")
sigGenes <- pathwaysGSE$summary |> dplyr::filter(Input == "sig")
if (nrow(sigGenes) == 0) {
  data <- data.frame()
} else {
  data <- table(sigGenes$DB) |>
    as.data.frame() |>
    tidyr::pivot_wider(names_from = Var1, values_from = Freq)
}
print(htmltools::tagList(DT::datatable(data, rownames = FALSE, style = "bootstrap4", options = list(dom = 't'))))

cat(paste0("\n## ", nameMap$overRep, "\n"))
cat(descriptionMap$overRep, "\n")
cat("\n Only Significant genes as input:\n")
if (nrow(pathwaysOverRep$summary) == 0) {
  data <- data.frame()
} else {
  data <- table(pathwaysOverRep$summary$Input, pathwaysOverRep$summary$DB) |>
    as.data.frame() |>
    tidyr::pivot_wider(names_from = Var2, values_from = Freq) |>
    dplyr::rename("Input" = "Var1")
}
print(htmltools::tagList(DT::datatable(data, rownames = FALSE, style = "bootstrap4", options = list(dom = 't'))))
```

::: {.callout-note collapse="true"}
## Session information
```{r sessionInfo}
sessionInfo()
```
:::