# Summary

Here is a summary of the results. I compile the results from the different comparisons in one table.

```{r setup}
#| warning: FALSE
#| message: FALSE
#| error: false
set.seed(42)
# load necessary packages
libraries <- c(
  "DESeq2", "edgeR", "ggplot2", "dplyr", "tximport", "limma", "ggVennDiagram", "cowplot",
  "openxlsx", "DT", "org.Mm.eg.db", "clusterProfiler", "RColorBrewer", "pheatmap", "ReactomePA", "msigdbr"
)
loaded <- lapply(libraries, require, character.only = TRUE)

projectDir <- getwd()
datasetResDir <- file.path(projectDir, "results")

# source all files in the R directory
invisible(lapply(list.files(file.path(projectDir, "R"), full.names = TRUE), source))

padj.cutoff <- 0.05
lfc.cutoff <- 0.6
```

```{r openxlsxStyles}
#| echo: false
#| warning: false
#| message: false
byExpStyle <- openxlsx::createStyle(fgFill = "#e0edd3", borderColour = "black", border = c("left", "right"), textRotation = 45, textDecoration = "bold", fontSize = 14)
bySexRAStyle <- openxlsx::createStyle(fgFill = "#f5e8d1", borderColour = "black", border = c("left", "right"), textRotation = 45, textDecoration = "bold", fontSize = 14)
femaleStyle <- openxlsx::createStyle(fgFill = "#f0d9f4", borderColour = "black", border = c("left"), textRotation = 45, textDecoration = "bold", fontSize = 14)
maleStyle <- openxlsx::createStyle(fgFill = "#dbf7f7", borderColour = "black", border = c("right"), textRotation = 45, textDecoration = "bold", fontSize = 14)
sidesBorderStyle <- openxlsx::createStyle(borderColour = "black", border = c("left", "right"))
leftBorderStyle <- openxlsx::createStyle(borderColour = "black", border = c("left"))
rightBorderStyle <- openxlsx::createStyle(borderColour = "black", border = c("right"))
```

## Differential Expression

Generated an excel file of all the DESeq2 DE results at: `r file.path(datasetResDir, "Differential_Expression_Summary.xlsx")`

```{r}
#| echo: false
#| warning: false
#| message: false
byExpDE <- readRDS(file.path(datasetResDir, "byExp", "objects", "DE.rds"))$DESeq2 |> dplyr::filter(shrunken_sig_exp != "NO")
male_byExpSexDE <- readRDS(file.path(datasetResDir, "byExpSex", "objects", "Male_DE.rds"))$DESeq2 |> dplyr::filter(shrunken_sig_exp != "NO")
female_byExpSexDE <- readRDS(file.path(datasetResDir, "byExpSex", "objects", "Female_DE.rds"))$DESeq2 |> dplyr::filter(shrunken_sig_exp != "NO")
bySexRA <- readRDS(file.path(datasetResDir, "bySexRA", "objects", "DE.rds"))$DESeq2 |> dplyr::filter(shrunken_sig_exp != "NO")

summary <- merge_nested_df(list(HOvsRA = byExpDE, Female_HOvsRA = female_byExpSexDE, Male_HOvsRA = male_byExpSexDE, RA_MvsF = bySexRA)) |>
  dplyr::rename(Comparison = Level1) |>
  dplyr::relocate(Comparison, .before = "baseMean") |>
  dplyr::rename(ENSEMBL = row) |>
  dplyr::relocate(SYMBOL, .after = "ENSEMBL") |>
  dplyr::select(-sig_exp, -lfcSE, -log2FoldChange)

wb <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb, "summary")
openxlsx::addWorksheet(wb, "Oxygen vs Room Air")
openxlsx::writeData(wb, "Oxygen vs Room Air", summary |> dplyr::filter(Comparison == "HOvsRA") |> dplyr::select(-Comparison) |> dplyr::arrange(desc(shrunken_log2FoldChange)))
openxlsx::addWorksheet(wb, "Female - Oxygen vs Room Air")
openxlsx::writeData(wb, "Female - Oxygen vs Room Air", summary |> dplyr::filter(Comparison == "Female_HOvsRA") |> dplyr::select(-Comparison) |> dplyr::arrange(desc(shrunken_log2FoldChange)))
openxlsx::addWorksheet(wb, "Male - Oxygen vs Room Air")
openxlsx::writeData(wb, "Male - Oxygen vs Room Air", summary |> dplyr::filter(Comparison == "Male_HOvsRA") |> dplyr::select(-Comparison) |> dplyr::arrange(desc(shrunken_log2FoldChange)))
openxlsx::addWorksheet(wb, "Male vs Female - RA")
openxlsx::writeData(wb, "Male vs Female - RA", summary |> dplyr::filter(Comparison == "RA_MvsF") |> dplyr::select(-Comparison) |> dplyr::arrange(desc(shrunken_log2FoldChange)))

summary <- summary |> 
  tidyr::pivot_wider(names_from = Comparison, values_from = -c(ENSEMBL, SYMBOL, Comparison), names_glue = "{Comparison}_{.value}") |>
  dplyr::relocate(contains("shrunken_log2FoldChange"), contains("padj"), contains("shrunken_sig_exp"), contains("pvalue"), contains("stat"), contains("baseMean"), contains("shrunken_lfcSE"), .after = "SYMBOL") |>
  dplyr::rename_at(dplyr::vars(contains("shrunken_log2FoldChange")), ~stringr::str_replace_all(.x, "shrunken_log2FoldChange", "log2FC"))
openxlsx::writeData(wb, "summary", summary)

openxlsx::addStyle(wb, sheet = "summary", style = byExpStyle, rows = 1, cols = grep("HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = bySexRAStyle, rows = 1, cols = grep("RA_MvsF_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = maleStyle, rows = 1, cols = grep("Male_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = femaleStyle, rows = 1, cols = grep("Female_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = sidesBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = sidesBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("RA_MvsF_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = rightBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("Male_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = leftBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("Female_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::freezePane(wb, "summary", firstRow = TRUE, firstActiveCol = 3, firstCol = TRUE)
    openxlsx::setColWidths(wb, "summary", cols = 1, widths = "auto")
openxlsx::saveWorkbook(wb, file.path(datasetResDir, "Differential_Expression_Summary.xlsx"), overwrite = TRUE)
```

## Pathway Analysis

Generated an excel spreadsheet with all the GSEA and Over Representation pathway analysis results at:

- `r file.path(datasetResDir, "GSEA_summary.xlsx")`
- `r file.path(datasetResDir, "OverRep_summary.xlsx")`

```{r GSEA}
#| echo: false
#| warning: false
#| message: false
byExpGSE <- readRDS(file.path(datasetResDir, "byExp", "objects", "pathwayListGSE.rds"))$summary |> dplyr::select(-row)
male_byExpSexGSE <- readRDS(file.path(datasetResDir, "byExpSex", "objects", "Male_pathwayListGSE.rds"))$summary |> dplyr::select(-row)
female_byExpSexGSE <- readRDS(file.path(datasetResDir, "byExpSex", "objects", "Female_pathwayListGSE.rds"))$summary |> dplyr::select(-row)
bySexRA <- readRDS(file.path(datasetResDir, "bySexRA", "objects", "pathwayListGSE.rds"))$summary |> dplyr::select(-row)

summary <- merge_nested_df(list(HOvsRA = byExpGSE, Male_HOvsRA = male_byExpSexGSE, Female_HOvsRA = female_byExpSexGSE, RA_MvsF = bySexRA)) |>
  dplyr::rename(Comparison = Level1) |>
  dplyr::relocate(Comparison, .before = "Input") |>
  dplyr::select(-row)

wb <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb, "summary")

openxlsx::addWorksheet(wb, "Oxygen vs Room Air")
openxlsx::writeData(wb, "Oxygen vs Room Air", summary |> dplyr::filter(Comparison == "HOvsRA") |> dplyr::select(-Comparison) |> dplyr::arrange(desc(NES)))
openxlsx::addWorksheet(wb, "Female - Oxygen vs Room Air")
openxlsx::writeData(wb, "Female - Oxygen vs Room Air", summary |> dplyr::filter(Comparison == "Female_HOvsRA") |> dplyr::select(-Comparison) |> dplyr::arrange(desc(NES)))
openxlsx::addWorksheet(wb, "Male - Oxygen vs Room Air")
openxlsx::writeData(wb, "Male - Oxygen vs Room Air", summary |> dplyr::filter(Comparison == "Male_HOvsRA") |> dplyr::select(-Comparison) |> dplyr::arrange(desc(NES)))
openxlsx::addWorksheet(wb, "Male vs Female - RA")
openxlsx::writeData(wb, "Male vs Female - RA", summary |> dplyr::filter(Comparison == "RA_MvsF") |> dplyr::select(-Comparison) |> dplyr::arrange(desc(NES)))

summary <- tidyr::pivot_wider(summary, names_from = Comparison, values_from = -c("DB", "Comparison", "Input", "ID", "Description"), names_glue = "{Comparison}_{.value}") |>
  dplyr::relocate(contains("NES"), contains("p.adjust"), contains("setSize"), contains("core_enrichment"), .after = "Description") 

openxlsx::writeData(wb, "summary", summary)

openxlsx::addStyle(wb, sheet = "summary", style = byExpStyle, rows = 1, cols = grep("HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = bySexRAStyle, rows = 1, cols = grep("RA_MvsF_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = maleStyle, rows = 1, cols = grep("Male_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = femaleStyle, rows = 1, cols = grep("Female_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = sidesBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = sidesBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("RA_MvsF_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = rightBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("Male_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = leftBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("Female_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::freezePane(wb, "summary", firstRow = TRUE, firstCol = FALSE)
openxlsx::setColWidths(wb, "summary", cols = 1, widths = "auto")

openxlsx::saveWorkbook(wb, file = file.path(datasetResDir, "GSEA_summary.xlsx"), overwrite = TRUE)
```

```{r OverRep}
#| echo: false
#| warning: false
#| message: false
byExpOverRep <- readRDS(file.path(datasetResDir, "byExp", "objects", "pathwayListOverRep.rds"))$summary |> dplyr::select(-row)
male_byExpSexOverRep <- readRDS(file.path(datasetResDir, "byExpSex", "objects", "Male_pathwayListOverRep.rds"))$summary |> dplyr::select(-row)
female_byExpSexOverRep <- readRDS(file.path(datasetResDir, "byExpSex", "objects", "Female_pathwayListOverRep.rds"))$summary |> dplyr::select(-row)
bySexRAOverRep <- readRDS(file.path(datasetResDir, "bySexRA", "objects", "pathwayListOverRep.rds"))$summary |> dplyr::select(-row)

summary <- merge_nested_df(list(HOvsRA = byExpOverRep, Male_HOvsRA = male_byExpSexOverRep, Female_HOvsRA = female_byExpSexOverRep, RA_MvsF = bySexRAOverRep)) |>
  dplyr::rename(Comparison = Level1) |>
  dplyr::relocate(Comparison, .before = "Input") |>
  dplyr::select(-row)

wb <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb, "summary")

openxlsx::addWorksheet(wb, "Oxygen vs Room Air")
openxlsx::writeData(wb, "Oxygen vs Room Air", summary |> dplyr::filter(Comparison == "HOvsRA") |> dplyr::select(-Comparison) |> dplyr::arrange(p.adjust))
openxlsx::addWorksheet(wb, "Female - Oxygen vs Room Air")
openxlsx::writeData(wb, "Female - Oxygen vs Room Air", summary |> dplyr::filter(Comparison == "Female_HOvsRA") |> dplyr::select(-Comparison) |> dplyr::arrange(p.adjust))
openxlsx::addWorksheet(wb, "Male - Oxygen vs Room Air")
openxlsx::writeData(wb, "Male - Oxygen vs Room Air", summary |> dplyr::filter(Comparison == "Male_HOvsRA") |> dplyr::select(-Comparison) |> dplyr::arrange(p.adjust))
openxlsx::addWorksheet(wb, "Male vs Female - RA")
openxlsx::writeData(wb, "Male vs Female - RA", summary |> dplyr::filter(Comparison == "RA_MvsF") |> dplyr::select(-Comparison) |> dplyr::arrange(p.adjust))

summary <- tidyr::pivot_wider(summary, names_from = Comparison, values_from = -c("DB", "Comparison", "Input", "ID", "Description"), names_glue = "{Comparison}_{.value}") |>
  dplyr::relocate(contains("p.adjust"), contains("GeneRatio"), contains("geneID"), .after = "Description") 

openxlsx::writeData(wb, "summary", summary)

openxlsx::addStyle(wb, sheet = "summary", style = byExpStyle, rows = 1, cols = grep("HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = bySexRAStyle, rows = 1, cols = grep("RA_MvsF_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = maleStyle, rows = 1, cols = grep("Male_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = femaleStyle, rows = 1, cols = grep("Female_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = sidesBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = sidesBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("RA_MvsF_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = rightBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("Male_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::addStyle(wb, sheet = "summary", style = leftBorderStyle, rows = 2:(nrow(summary) + 1), cols = grep("Female_HOvsRA_", colnames(summary)), gridExpand = TRUE)
openxlsx::freezePane(wb, "summary", firstRow = TRUE, firstCol = FALSE)
openxlsx::setColWidths(wb, "summary", cols = 1, widths = "auto")

openxlsx::saveWorkbook(wb, file = file.path(datasetResDir, "OverRep_summary.xlsx"), overwrite = TRUE)
```

::: {.callout-note collapse="true"}
## Session information
```{r sessionInfo}
sessionInfo()
```
:::